---
title: "Biostat 213 Homework 1"
author: Zian ZHUANG
output:
  html_document:
    highlight: pygments
    toc: yes
    toc_float: yes
---

<!-- Setup -->
<style type="text/css">
body{ /* Normal  */
      font-size: 17px;
      font-family: "Times New Roman";
  }
h1,h2,h3,h4,h5,h6{
  font-family: "Times New Roman";
}
</style>
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, cache.lazy = FALSE)
library(kableExtra)
library(tidyverse)
library(gt)
```
<!-- Begin writing -->
## Bivariate Gibbs

Generate 2 normal distributions X and Y with known correlation $\rho$.

### set up function 

```{r}
set.seed(1996)
mcmc_bi <- function(rho){ # input correlation
  m = 100000; x = y = numeric(m); x[1] = y[1] = 0  # initial conditions
  sgm = sqrt(1 - rho^2)                            # standard deviation
  
  for (i in 2:m) {                                 # Gibbs Sampler Loop
    x[i] = rnorm(1, rho*y[i-1], sgm)
    y[i] = rnorm(1, rho*x[i], sgm)
  }
  
  alpha.1 = beta.1 = .01
  alpha.2 = .015; beta.2 = .012
  x1 = alpha.1 +beta.1*x          # x1 ~ N(alpha.1, beta.1)
  y1 = alpha.2 +beta.2*y          # x2 ~ N(alpha.2, beta.2)
  a = .4; b = .6                    #  weights
  w = a*x1 + b*y1; mean(w); sd(w) # Overall returns
  return(list(x1=x1, y1=y1, x=x, y=y, w=w))
}
```

### $\rho=0.1$

```{r}
sim1 <- mcmc_bi(rho = 0.1)
par(mfrow=c(2,3), pty="s") # 2 x 2 array of square plots
plot(sim1$x[1:50], sim1$y[1:50], type="l")
plot(sim1$x1, sim1$y1, pch=".", 
     main = "x1, x2 returns, green = portfolio > 0")
points(sim1$x1[sim1$w>0], sim1$y1[sim1$w>0], pch=".", col="green4")
plot(sim1$x[1:1000], type="l"); plot(sim1$y[1:1000], type="l")
acf(sim1$y); acf(sim1$x)

```

### $\rho=0.5$

```{r}
sim2 <- mcmc_bi(rho = 0.5)
par(mfrow=c(2,3), pty="s") # 2 x 2 array of square plots
plot(sim2$x[1:50], sim2$y[1:50], type="l")
plot(sim2$x1, sim2$y1, pch=".", 
     main = "x1, x2 returns, green = portfolio > 0")
points(sim2$x1[sim2$w>0], sim2$y1[sim2$w>0], pch=".", col="green4")
plot(sim2$x[1:1000], type="l"); plot(sim2$y[1:1000], type="l")
acf(sim2$y);acf(sim2$x)
```

### $\rho=0.7$

```{r}
sim3 <- mcmc_bi(rho = 0.7)
par(mfrow=c(2,3), pty="s") # 2 x 2 array of square plots
plot(sim3$x[1:50], sim3$y[1:50], type="l")
plot(sim3$x1, sim3$y1, pch=".", 
     main = "x1, x2 returns, green = portfolio > 0")
points(sim3$x1[sim3$w>0], sim3$y1[sim3$w>0], pch=".", col="green4")
plot(sim3$x[1:1000], type="l"); plot(sim3$y[1:1000], type="l")
acf(sim3$y);acf(sim3$x)
```

### $\rho=0.99$

```{r}
sim4 <- mcmc_bi(rho = 0.95)
par(mfrow=c(2,3), pty="s") # 2 x 2 array of square plots
plot(sim4$x[1:50], sim4$y[1:50], type="l")
plot(sim4$x1, sim4$y1, pch=".", 
     main = "x1, x2 returns, green = portfolio > 0")
points(sim4$x1[sim4$w>0], sim4$y1[sim4$w>0], pch=".", col="green4")
plot(sim4$x[1:1000], type="l"); plot(sim4$y[1:1000], type="l")
acf(sim4$y); acf(sim4$x)
```

### $\rho=1$

```{r}
sim5 <- mcmc_bi(rho = 1)
par(mfrow=c(2,2), pty="s") # 2 x 2 array of square plots
plot(sim5$x[1:50], sim5$y[1:50], type="l")
plot(sim5$x1, sim5$y1, pch=".", 
     main = "x1, x2 returns, green = portfolio > 0")
points(sim5$x1[sim5$w>0], sim5$y1[sim5$w>0], pch=".", col="green4")
plot(sim5$x[1:1000], type="l"); plot(sim5$y[1:1000], type="l")
```

Since x and y are unchanged (sd=0) in the chain, we cannot calculate the ACF plots for them.

### Conclusion

When correlation between x and y increase (i.e., $\rho$ increases), it is harder for chains to converge. As we can tell from the ACF plots above, When correlation increase, it will approach to 0 more slowly. In addition, plots of x and y shows that, when correlation is large (e.g., $\rho=0.99$), points are wandering around, showing the poor convergence. $\rho=1$ is a special case since the standard deviation equals 0 and x, y take values from each other. In this case, x, y are unchanged within the chain.

```{r}
x1_total <- cbind(sim1$x1, sim2$x1, sim3$x1, sim4$x1, sim5$x1)
df <- apply(x1_total, 2, 
            function(x)quantile(x, c(0.025,0.5,0.975))) %>% round(.,4)
colnames(df) <- c("rho=0.1","rho=0.5","rho=0.7","rho=0.99","rho=0.1")
df %>%
  kbl(caption = "Summary table of x1") %>%
  kable_classic(full_width = F, html_font = "Cambria")

y1_total <- cbind(sim1$y1, sim2$y1, sim3$y1, sim4$y1, sim5$y1)
df <- apply(y1_total, 2, 
            function(x)quantile(x, c(0.025,0.5,0.975))) %>% round(.,4)
colnames(df) <- c("rho=0.1","rho=0.5","rho=0.7","rho=0.99","rho=1")
df %>%
  kbl(caption = "Summary table of y1") %>%
  kable_classic(full_width = F, html_font = "Cambria")
```

From the summary table of x1 and y1, we can see that all values of rho provides reasonable estimates, except for rho=1. Then we know that if the chain is long enough, we may still be able to obtain valid estimates of highly correlated variables.

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>